\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2019-01-31 Documents/M1/Astronum_2018/sections/TimeIntegration.tex

Since the set of realizable moments is convex, convex-invariant schemes, which preserve the initial value's convex domain, can be used to design realizability-preserving methods for the two-moment model. 
Ideally, the scheme should also be high-order accurate and work well in the asymptotic diffusion limit (characterized by frequent collisions and long time scales).  
The following discussion considers the construction of such convex-invariant schemes.

@@

\subsection{Convex-Invariant Implicit-Explicit Schemes}

To be convex-invariant, the coefficients and weights defining the IMEX scheme must satisfy additional constraints.
Our goal is to find constraints on $a_{ij}$, $\tilde{a}_{ij}$, $\tilde{w}_{i}$, and $w_{i}$ that enable each $\vect{u}^{(i)}$ in Eq.~\eqref{imexStages} to be expressed as a convex combination of realizable states.  
Following Hu et al.~\cite{hu_etal_2018}, the stage values in Eq.~\eqref{imexStages} can be rewritten as
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big] + a_{ii}\,\dt\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$ and $\hat{c}_{ij}\equiv\tilde{c}_{ij}/c_{ij}$ are defined in terms of $a_{ij}$ and $\tilde{a}_{ij}$.
For IMEX schemes of type~ARS, $c_{ij}$ and $\tilde{c}_{ij}$ are given by
    \begin{equation}
     \begin{aligned}
      c_{i0} &= 1-\sum_{j=2}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
      c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj}, \\
      \tilde{c}_{i0} &= \tilde{a}_{i1}+\sum_{j=2}^{i-1}a_{ij}\tilde{b}_{j1}, \quad &
      \tilde{c}_{ij} &= \tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj},  
     \end{aligned}
     \label{eq:positivityCoefficientsARS}
    \end{equation}
    \begin{equation}
      b_{ii} = \f{1}{a_{ii}}, \quad
      b_{ij} = -\f{1}{a_{ii}}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad
      \tilde{b}_{ij} = -\f{1}{a_{ii}}\Big(\tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}\Big).  
    \end{equation}
Note that $c_{i1}=\tilde{c}_{i1}=0$ in Eq.~\eqref{eq:positivityCoefficientsARS}, so that $\sum_{j=0}^{i-1}c_{ij}=1$.

If the IMEX scheme is GSA, $\vect{u}^{n+1} = \vect{u}^{(s)}$.  
Moreover, if $c_{ij},\tilde{c}_{ij}\ge0$ and $a_{ii}>0$, each stage in Eq.~\eqref{eq:imexStagesRewrite} is a convex combination of explicit Euler steps (with time step $\hat{c}_{ij}\dt$), followed by a implicit Euler step.  
Each of the explicit Euler steps has a time step condition that ensures its realizability given by $\hat{c}_{ij}\,\dt\leq\dx$; the Courant-Friedrichs-Lewy (CFL) condition of the scheme.
Using results proved in~\cite{chu_etal_2018} and discussed in Section~\ref{se:SpacialDiscretization}, the IMEX scheme is convex-invariant and realizability-preserving for the two-moment model in Section~\ref{se:SpacialDiscretization} provided
\begin{equation}
  \max(\hat{c}_{ij})\,\dt \leq \dx.  
\end{equation}
(This CFL condition becomes more restrictive with high-order DG spatial discretization~\cite{chu_etal_2018}.)

\subsection{Diffusion Accurate, Convex-Invariant Implicit-Explicit Schemes}

Accuracy in the diffusion limit is another important property to consider for the IMEX scheme applied to the two-moment model.  
In the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$, and the two-moment model is approximately governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
In the context of IMEX schemes, the above relationships imply that the following relations should hold~\cite{chu_etal_2018}
\begin{equation}
   \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=1,\ldots,s,
   \label{diffusionAccuracy}
\end{equation}
where $\vect{e}_{i}$ is the $i$th column of the $s\times s$ identity matrix, $\vect{e}$ is the vector of ones, and $A$ and $\tilde{A}$ are the matrices of the coefficients $(\tilde{a}_{ij})$ and $(a_{ij})$.
Eq.~\eqref{diffusionAccuracy} implies
\begin{equation}
  c_{i} = \tilde{c}_{i}, \quad i=1,\ldots,s
\end{equation}
We have proved in \cite{chu_etal_2018} that only IMEX schemes of type~ARS can be both diffusion accurate and convex-invariant.  
(Another short proof follows from the fact that IMEX schemes of type~A have $\tilde{c}_1 = 0$ while $c_i \neq 0$.)

\subsection{PD-ARS schemes}

Unfortunately, coefficients satisfying the order conditions in Eqs.~\eqref{orderConditions1}-\eqref{orderConditions2} and the conditions for convex-invariance do not exist for the standard IMEX scheme in Eqs.~\eqref{imexStages}-\eqref{imexIntermediate}, unless a small time step is invoked that makes the scheme essentially explicit.  
To circumvent this problem, correction steps can be introduced after the assembly step in Eq.~\eqref{imexIntermediate} (e.g., \cite{chertock_etal_2015,hu_etal_2018}).  
However, the correction steps can impose time step constraints for realizability or accuracy in the diffusion limit that ruins the efficiency gains expected from the IMEX scheme.  
Because of this, we sacrifice overall high-order accuracy, and aim for IMEX schemes that are high-order accurate in the streaming limit, diffusion accurate, and convex-invariant.  
Combining these requirements we seek GSA IMEX schemes of type~ARS with coefficients satisfying the following constraints~\cite{hu_etal_2018,chu_etal_2018}:
\begin{enumerate}
    \item Consistency of the implicit coefficients:
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item High-order accuracy in the streaming limit.
    For second-order accuracy:
    \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2}.
      \label{eq:orderConditionsEx}
    \end{equation}
    For third-order accuracy: 
    \begin{equation}
    \sum_{i=1}^{s}\tilde{w}_{i}=1,
          \quad
          \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
          \quad
          \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c_{i}}^2 = \f{1}{3}
          \quad\text{and}\quad
          \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{a_{ij}}\tilde{c}_{j}= \f{1}{6}
    \end{equation}
    \item Diffusion accurate:
    \begin{equation}
      c_{i}=\tilde{c}_{i}, \quad i=1,\ldots,s.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Convex-invariant:
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
      \label{eq:convexInvariant}
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}:=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.
    \item Less than five stages ($s\le4$) \label{cod:statges}.
    \item Globally stiffly accurate: $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$. 
\end{enumerate}
Fortunately, these IMEX schemes are easy to find.  
(The constraint in \eqref{cod:statges} is introduced from efficiency considerations to limit the number of implicit solves.)
We call the IMEX schemes satisfying the above conditions {PD-ARS} (see also Definition~3 in~\cite{chu_etal_2018}), and we provide two optimal PD-ARS schemes below: (PD-ARS with SSPRK2) and (PD-ARS with SSPRK3), which limit to the optimal second-order and third-order SSPRK schemes from~\cite{shuOsher_1988}, respectively.

\subsubsection{PD-ARS with SSPRK2}

The optimal PD-ARS with SSPRK2 in the standard double Butcher tableau form, explicit tableau on the left ($\tilde{A}$) and implicit tableau on the right ($A$), is given by:
\begin{align}
  &\begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0 
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2 & 1/2 \\ \hline
  	  & 0 & 1/2 & 1/2
  \end{array} \\ 
  &\hspace{48pt} \tilde{A} \hspace{90pt} A
\end{align}
For this scheme, only two implicit solves are needed per time step and $c_{\mbox{\tiny Sch}}= 1$, which implies that the time step restriction for preserving moment realizability is only due to the explicit part.

\subsubsection{PD-ARS with SSPRK3}

The optimal PD-ARS with SSPRK3 is given in its standard double Butcher tableau form (explicit tableau on the left and implicit tableau on the right) by:
\begin{align}
  &\begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6 & 1/6 & 2/3\\\hline
  	  & 0 & 1/6 & 1/6 & 2/3
  \end{array}\\ 
    &\hspace{60pt} \tilde{A} \hspace{130pt} A
\end{align}
This scheme requires three implicit solves per time step, and $c_{\mbox{\tiny Sch}}= 1$.  
Since (PD-ARS with SSPRK3) is not more accurate than (PD-ARS with SSPRK2) in collision dominated regions (see results in Section~\ref{se:NumericalTests}), it may not offer any practical advantage over (PD-ARS with SSPRK2).

>>>>

Since the set of realizable moments is convex, convex-invariant schemes, which maps the initial value remains in its domain, can be used to design realizability-preserving methods for the two-moment model.
Ideally, the scheme should also be high-order accurate and work well in the asymptotic diffusion limit (characterized by frequent collisions and long time scales).  
The following discussion considers the construction of such convex-invariant schemes.

@@

\subsection{Convex-Invariant IMEX Schemes}

To be convex-invariant, the coefficients and weights defining the IMEX scheme must satisfy additional constraints.
Our goal is to find constraints on $a_{ij}$, $\tilde{a}_{ij}$, $\tilde{w}_{i}$, and $w_{i}$ that enable each $\vect{u}^{(i)}$ in Eq.~\eqref{imexStages} to be expressed as a convex combination of realizable states.  
Following Hu et al.~\cite{hu_etal_2018}, the stage values in Eq.~\eqref{imexStages} can be rewritten as
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big] + a_{ii}\,\dt\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$ and $\hat{c}_{ij}\equiv\tilde{c}_{ij}/c_{ij}$ are defined in terms of $a_{ij}$ and $\tilde{a}_{ij}$.
For IMEX schemes of type~ARS, $c_{ij}$ and $\tilde{c}_{ij}$ are given by~\cite{hu_etal_2018}
    \begin{equation}
     \begin{aligned}
      c_{i0} &= 1-\sum_{j=2}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
      c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj}, \\
      \tilde{c}_{i0} &= \tilde{a}_{i1}+\sum_{j=2}^{i-1}a_{ij}\tilde{b}_{j1}, \quad &
      \tilde{c}_{ij} &= \tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj},  
     \end{aligned}
     \label{eq:positivityCoefficientsARS}
    \end{equation}
    \begin{equation}
      b_{ii} = \f{1}{a_{ii}}, \quad
      b_{ij} = -\f{1}{a_{ii}}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad
      \tilde{b}_{ij} = -\f{1}{a_{ii}}\Big(\tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}\Big).  
    \end{equation}
Note that $c_{i1}=\tilde{c}_{i1}=0$ in Eq.~\eqref{eq:positivityCoefficientsARS}, so that $\sum_{j=0}^{i-1}c_{ij}=1$.

If the IMEX scheme is GSA, $\vect{u}^{n+1} = \vect{u}^{(s)}$.  
Moreover, if $c_{ij},\tilde{c}_{ij}\ge0$ and $a_{ii}>0$, each stage in Eq.~\eqref{eq:imexStagesRewrite} is a convex combination of explicit Euler steps (with time step $\hat{c}_{ij}\dt$), followed by a implicit Euler step.  
Each of the explicit Euler steps has a time step condition that ensures its realizability given by $\hat{c}_{ij}\,\dt\leq\dx$; the CFL condition of the scheme.
Using results proved in~\cite{chu_etal_2018} and discussed in Section~\ref{se:SpatialDiscretization}, the IMEX scheme is convex-invariant and realizability-preserving for the two-moment model in Section~\ref{se:SpatialDiscretization} provided
\begin{equation}
  \max(\hat{c}_{ij})\,\dt \leq \dx.  
\end{equation}
(This CFL condition becomes more restrictive with high-order DG spatial discretization~\cite{chu_etal_2018}.)

\subsection{Diffusion Accurate, Convex-Invariant IMEX Schemes}

Accuracy in the diffusion limit is another important property to consider when an IMEX scheme is applied to the two-moment model.  
In the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$, and the two-moment model is approximately governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
In the context of IMEX schemes, the above relationships imply that the following relations should hold~\cite{chu_etal_2018}:
\begin{equation}
   \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=1,\ldots,s,
   \label{diffusionAccuracy}
\end{equation}
where $\vect{e}_{i}$ is the $i$th column of the $s\times s$ identity matrix, $\vect{e}$ is the vector of ones, and $A$ and $\tilde{A}$ are the matrices of the coefficients $(\tilde{a}_{ij})$ and $(a_{ij})$.
Eq.~\eqref{diffusionAccuracy} implies:
\begin{equation}
  c_{i} = \tilde{c}_{i}, \quad i=1,\ldots,s.
\end{equation}
We have proved in \cite{chu_etal_2018} that only IMEX schemes of type~ARS can be both diffusion accurate and convex-invariant.  
(Another short proof follows from the fact that IMEX schemes of type~A have $\tilde{c}_1 = 0$ while $c_i \neq 0$.)

\subsection{PD-ARS IMEX schemes}

Unfortunately, coefficients satisfying the order conditions in Eqs.~\eqref{orderConditions1}-\eqref{orderConditions2} and the conditions for convex-invariance do not exist for the standard IMEX scheme in Eqs.~\eqref{imexStages}-\eqref{imexIntermediate}, unless a small time step is invoked that makes the scheme essentially explicit.  
To circumvent this problem, correction steps can be introduced after the assembly step in Eq.~\eqref{imexIntermediate} (e.g., \cite{chertock_etal_2015,hu_etal_2018}).  
However, the correction steps can impose time step constraints for realizability or accuracy in the diffusion limit that ruin the efficiency gains expected from the IMEX scheme.  
Because of this, we sacrifice overall high-order accuracy, and aim for IMEX schemes that are high-order accurate in the streaming limit, diffusion accurate, and convex-invariant.  
Combining these requirements we seek GSA IMEX schemes of type~ARS with coefficients satisfying the following constraints~\cite{hu_etal_2018,chu_etal_2018}:
\begin{enumerate}
    \item Consistency of the implicit coefficients:
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item High-order accuracy in the streaming limit.
    For second-order accuracy:
    \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2}.
      \label{eq:orderConditionsEx}
    \end{equation}
    For third-order accuracy: 
    \begin{equation}
    \sum_{i=1}^{s}\tilde{w}_{i}=1,
          \quad
          \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
          \quad
          \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c_{i}}^2 = \f{1}{3}
          \quad\text{and}\quad
          \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{a_{ij}}\tilde{c}_{j}= \f{1}{6}.
    \end{equation}
    \item Diffusion accuracy:
    \begin{equation}
      c_{i}=\tilde{c}_{i}, \quad i=1,\ldots,s.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Convex-invariance:
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1,
      \label{eq:convexInvariant}
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}:=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.

    (Note that the greater the $c_{\Sch}$, the larger the time step can be.
    And $c_{\Sch} \leq 1$.)
    \item Have less than five stages ($s\le4$)\label{cod:statges}.
    \item Are globally stiffly accurate: $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$. 
\end{enumerate}
Fortunately, these IMEX schemes are easy to find.  
(The constraint in \eqref{cod:statges} is introduced from efficiency considerations to limit the number of implicit solves.)
We call the IMEX schemes satisfying the above conditions {PD-ARS} (see also Definition~3 in~\cite{chu_etal_2018}), and we provide two optimal PD-ARS schemes below: PD-ARS2 and PD-ARS3, each limits to the optimal second-order and third-order SSPRK schemes from~\cite{shuOsher_1988}, respectively.
\subsubsection{PD-ARS2}

The optimal 3-statge PD-ARS, PD-ARS2, in the standard double Butcher tableau form, with explicit tableau ($\tilde{A}$) on the left and implicit tableau ($A$) on the right, is given by
\begin{align}
  &\begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0 
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2 & 1/2 \\ \hline
  	  & 0 & 1/2 & 1/2
  \end{array}
\end{align}
Note its explicit tableau is SSPRK2. 
For this scheme, only two implicit solves are needed per time step and $c_{\Sch}= 1$, which implies that the time step restriction for preserving moment realizability is only due to the explicit part.

\subsubsection{PD-ARS3}

The optimal 4-stage PD-ARS, PD-ARS3, is given in its standard double Butcher tableau form (explicit tableau on the left and implicit tableau on the right) by
\begin{align}
  &\begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6 & 1/6 & 2/3\\\hline
  	  & 0 & 1/6 & 1/6 & 2/3
  \end{array}
\end{align}
Its explicit tableau is SSPRK3. 
This scheme requires three implicit solves per time step, and $c_{\Sch}= 1$.  
Since PD-ARS3 is not more accurate than PD-ARS2 in collision dominated regions (see our results in Section~\ref{se:NumericalTests}), it may not offer any practical advantage over PD-ARS2.
